Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-22 _tex/index.tex
Built site for gh-pages

However, since the discriminal process of a single stimulus is not
directly observable, the \emph{law of comparative judgment} becomes
essential. This law states that in pairwise comparisons, the stimulus
positioned further along the trait continuum is perceived as having a
higher level of that trait. Thus, the theory assumes the observed
dichotomous outcome is determined by the distribution of the difference
between the underlying discriminal processes of the stimuli, referred to
as the \emph{discriminal difference}. This suggests that the outcome
depends on the relative distance between stimuli, rather than their
absolute positions on the trait continuum.

These concepts can be illustrated through an example. For instance, in
the context of evaluating text quality,
Figure~\ref{fig-discriminal_process} could depict the underlying
discriminal process distributions for two written texts, highlighting
differences in their discriminal dispersions and modal discriminal
processes along the quality trait continuum. Furthermore,
Figure~\ref{fig-discriminal_difference} could display the discriminal
difference distribution for these texts, showing that text A is
perceived to exhibit significantly higher quality than text B, as
indicated by the shaded gray area. Consequently, the dichotomous outcome
of this comparison would favor text A.

@@

Despite its reliance on the largest number of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity is largely due to its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one key difference: while Case V assumes a
Normal distribution for the discriminal processes of stimuli, the BTL
model uses the more mathematically tractable Logistic distribution
\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions differ by a scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I} (see
Figure~\ref{fig-logistic_vs_normal}).

@@

The assumption of \emph{zero correlation between stimuli} is more easily
understood through an example. For instance, when using pairwise
comparisons to evaluate text quality, the assumption suggests that a
judge's perception of the trait in one text does not influence his
perception of the same trait on another. Thurstone attributed this
independence to the cancellation of potential judges' biases, driven by
two opposing and equally weighted effects occurring during the pairwise
comparisons \citep[pp.~268]{Thurstone_1927b}. This cancellation was
mathematically demonstrated by \citet{Andrich_1978}, assuming
discriminal processes with additive biases. However, it is easy to
imagine two scenarios where the zero correlation assumption almost
certainly does not hold: when the pairwise comparison involves
multidimensional, complex traits with heterogeneous stimuli, and when an
additional hierarchical structure is relevant to the stimuli.

In the first scenario, evidence suggests that when judges are confronted
with multi-dimensional, complex traits and heterogeneous stimuli, such
as when evaluating written text quality, their assessments are
influenced by multiple, intricate aspects of the stimuli
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. In
such cases, it is not inconceivable that these various aspects of the
stimuli unevenly influence the judges' perception of one text over
another, leading to biases that resist cancellation, as the effects of
the factors involved in the comparison are neither equally weighted nor
opposing. For example, this may occur when a judge, while assessing the
argumentative quality of a text, places disproportionate emphasis on
grammatical errors, ultimately favoring texts with fewer errors but
weaker arguments. Although the literature does not provide direct
evidence for this specific argument, studies such as
\citet{Pollitt_et_al_2003} do offer evidence of judges' biases,
supporting the idea that the factors involved in pairwise comparisons
may not always cancel out.

@@

Building on the previous section, it is evident that the BTL model
commonly functions as the measurement model for the trait of interest in
CJ experiments \citep{Andrich_1978, Bramley_2008}. A measurement model
specifies how manifest variables contribute to the estimation of latent
variables \citep{Everitt_et_al_2010}. For example, when evaluating text
quality, researchers use the BTL model to process the dichotomous
outcomes resulting from the pairwise comparisons (the manifest
variables) to estimate scores that reflect the underlying quality level
of texts (the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

However, the statistical literature advises caution when using estimated
scores to conduct additional analyses and tests while warning against
eliminating data through ad-hoc univariate procedures. A key
consideration is that BTL scores are parameter estimates that inherently
carry uncertainty. Ignoring this uncertainty can bias the analysis and
reduce the precision of hypothesis tests. Notably, the direction and
magnitude of such biases are often unpredictable. Results may be
attenuated, exaggerated, or remain unaffected depending on the degree of
uncertainty in the scores and the actual effects being tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Moreover, eliminating data using ad-hoc univariate procedures using
these scores compounds this issue by excluding possible valuable
information from the analyses \citep{McElreath_2020}, further
introducing bias. Finally, the reduced precision in hypothesis tests
weakens their statistical power, increasing the likelihood of type-I or
type-II errors \citep{McElreath_2020}.

@@

\subsection{Assessment design features and their role on reliability and
validity}\label{sec-theory-issue3}

>>>>

However, since the discriminal process of a single stimulus is not
directly observable, the \emph{law of comparative judgment} becomes
essential. This law states that in pairwise comparisons, the stimulus
positioned further along the trait continuum is perceived as having a
higher level of that trait. Thus, the theory assumes the observed
dichotomous outcome is determined by the distribution of the difference
between the underlying discriminal processes of the stimuli, referred to
as the \emph{discriminal difference}. This indicates that the outcome
depends on the relative distance between stimuli, rather than their
absolute positions on the trait continuum.

These concepts are more easily understood through an example. For
instance, in the context of evaluating text quality,
Figure~\ref{fig-discriminal_process} could depict the underlying
discriminal process distributions for two written texts, highlighting
differences in their discriminal dispersions and modal discriminal
processes along the quality trait continuum. Furthermore,
Figure~\ref{fig-discriminal_difference} could display the discriminal
difference distribution for these texts, showing that text A is
perceived to exhibit significantly higher quality than text B, as
indicated by the shaded gray area. Consequently, the dichotomous outcome
of this comparison would probably favor text A.

@@

Despite its reliance on the largest number of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity is largely due to its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one key difference: while Case V assumes a
Normal distribution for the discriminal processes of stimuli, the BTL
model uses the more mathematically tractable Logistic distribution
\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions share similar statistical properties, differing only by a
scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I} (see
Figure~\ref{fig-logistic_vs_normal}).

@@

The assumption of \emph{zero correlation between stimuli} can be better
understood through an example. For instance, when using pairwise
comparisons to evaluate text quality, the assumption implies that a
judge's perception of a trait in one text does not influence his
perception of the same trait in another text. Thurstone attributed this
independence to the cancellation of potential judges' biases, driven by
two opposing and equally weighted effects occurring during the pairwise
comparisons \citep[pp.~268]{Thurstone_1927b}. This cancellation was
mathematically demonstrated by \citet{Andrich_1978}, using the BTL model
under the assumption of additive biases in the discriminal processes.
However, it is easy to imagine at least two scenarios where the zero
correlation assumption almost certainly does not hold: when the pairwise
comparison involves multidimensional, complex traits with heterogeneous
stimuli, and when an additional hierarchical structure is relevant to
the stimuli.

In the first scenario, the intricate aspects of multidimensional,
complex traits and heterogeneous stimuli may introduce dependencies
between stimuli. Research on text quality suggests that when judges
evaluate these traits, they often rely on various intricate aspects of
the stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. In
this context, it is not inconceivable that these aspects, being neither
equally weighted nor opposing, may unevenly influence judges'
perceptions, resulting in biases that resist cancellation. For example,
this might occur when a judge assessing the argumentative quality of a
text places disproportionate emphasis on grammatical accuracy,
ultimately favoring texts with fewer errors but weaker arguments. While
direct evidence for this specific scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of judges' biases,
supporting the idea that the different factors influencing pairwise
comparisons may not always cancel out.

In the second scenario, the shared context or inherent connections
created by the additional hierarchical structure may introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Nevertheless, despite recognizing
such hierarchical structures in CJ data, the statistical handling of
this extra source of dependency in the CJ literature has been
inadequate. For instance, in cases where the CJ data included multiple
samples of stimuli from the same individuals, researchers have often
relied on (averaged) estimated BTL scores to conduct subsequent analyses
and tests at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, has the significant limitation of ignoring the
uncertainty associated with the scores (refer to section
Section~\ref{sec-theory-issue2} for a detailed discussion of this
issue).

In contrast, the assumption of \emph{equal dispersion between stimuli}
suggests that the variability in the perceived trait level of the
stimuli is the same across stimuli. While Thurstone acknowledged that
this assumption may be violated when ``dealing with less conspicuous
attributes or with less homogeneous stimuli''
\citep[pp.~374]{Thurstone_1927a}, no study explicitly proposes that this
assumption could also be violated due to the presence of an additional
hierarchical (grouping) structure relevant to the texts. One such
scenario might arise, for example, when comparing texts produced by
university and secondary school students. In this case, university
students may consistently (or more precisely) produce higher-quality
texts, while secondary school students, who exhibit a broader range of
writing abilities, would show greater variability in the quality of
their texts. Although this example is somewhat contrived, it effectively
illustrates how assuming equal dispersions across texts can overlook
meaningful differences in the reliability of text quality across groups
or individuals.

{``however'' paragraph}

{``to mitigate'' paragraph}

@@

Building on the previous section, it is evident that the BTL model
commonly functions as the trait's measurement model in CJ experiments
\citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
manifest variables contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
researchers use the BTL model to process the dichotomous outcomes
resulting from the pairwise comparisons (the manifest variables) to
estimate scores that reflect the underlying quality level of texts (the
latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

However, the statistical literature advises caution when using estimated
scores for additional analyses and tests, as well as when eliminating
data through ad hoc univariate procedures. A key consideration is that
BTL scores are parameter estimates that inherently carry uncertainty.
Ignoring this uncertainty can bias the analysis and reduce the precision
of hypothesis tests. Notably, the direction and magnitude of such biases
are often unpredictable. Results may be attenuated, exaggerated, or
remain unaffected depending on the degree of uncertainty in the scores
and the actual effects being tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Moreover, excluding data using ad hoc univariate procedures can compound
these issues by discarding potentially valuable information, further
exacerbating the bias \citep{Zimmerman_1994, McElreath_2020}. Finally,
the reduced precision in hypothesis tests diminishes their statistical
power, increasing the likelihood of committing type-I or type-II errors
\citep{McElreath_2020}.

@@

\subsection{The diverse assessment design features and their role on
reliability and validity}\label{sec-theory-issue3}
